Deciding Bitvector Termination with Program Synthesis

### 2014-07-02 papers/termination/kalashnikov.tex
working on the kalashnikov section

In this section we present a solver for the synthesis fragment. An
important characteristic of our solver is that is models bit accurate
semantics. This enables the precise treatment of bit-wise operations and
arithmetic modulo fixed widths.

%We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
%have a program synthesiser kicking around, we present the design of one synthesiser
%that can be used to solve the problem.
%We start by recalling the synthesis formula:
% $$\exists P, x . \forall y . \sigma(P, x, y)$$
We start by recalling the synthesis fragment:
 \[
  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
 \]

@@

When the quantifier free first-order fragment of the logic is efficiently decidable and
$P$ is expressible as a ground term of the logic, we can use
Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
check the validity of the synthesis formula. The core of the CEGIS algorithm is
the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
detailed in Algorithm~\ref{fig:abstract-refinement-code}.

The algorithm is divided into two
components: {\sc synth} and {\sc verif}. For a finite set of test
inputs that we track, {\sc synth} synthesises a candidate program that is correct on just those tests.
Subsequently, {\sc Verif} tries to verify the candidate program by searching for an input on
which the candidate program does not satisfy the specification.  If no such input
can be found, the candidate program is correct.  Otherwise, the new input is added
to the setof tracked inputs.  
%If the input set is finite, this procedure is guaranteed to terminate.

@@

%\label{sec:logic}

The logic used to express our synthesis formula is \newC discussed in Section~\ref{sec:logic}.
%% The logic is a subset of C that we call \newC.  
%% The characteristic property of a
%% \newC  program is that safety can be decided for
%% it using a single query to a Bounded Model Checker.  A \newC program is
%% just a C program with the following syntactic restrictions:
%%  all loops in the program must have a constant bound;
%%  all recursion in the program must be limited to a constant depth;
%%  all arrays must be statically allocated (i.e. not using \texttt{malloc}),
%%  and be of constant size.
%% Additionally, \newC programs may use nondeterministic values, assumptions
%% and arbitrary-width types.

%% \iffalse
%% Two example \newC programs are shown 
%% in Fig.~\ref{fig:c-}.

%% \begin{figure}
%% \begin{minipage}[scale=0.8]{0.45\linewidth}
%%  \begin{lstlisting}[language=c]
%% int count_bits(int x) {
%%   int i, ret = 0;

%%   for (i = 0; i < 32; i++)
%%     if (x & (1 << i))
%%       ret++;

%%   return ret;
%% }
%%  \end{lstlisting}
%% \end{minipage}
%% \begin{minipage}{0.54\linewidth}
%%  \begin{lstlisting}[language=C]
%% int common_factor(int A[10]) {
%%   int i, factor = nondet();

%%   for (i = 0; i < 10; i++)
%%     assume((A[i] % factor) == 0);

%%   assume(factor > 1);
%%   return factor;
%% }

%%  \end{lstlisting}
%% \end{minipage}

%%  \caption{Two \newC programs}
%%  \label{fig:c-}

%% \end{figure}
%% \fi

%% Since each loop is bounded by a constant, and each recursive function call is
%% limited to a constant depth, a \newC program necessarily terminates and in
%% fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
%% a Bounded Model Checker with an unrolling bound of $k$ will be a complete
%% decision procedure for the safety of the program.  For a \newC program of
%% size $l$ and with largest loop bound~$k$, a Bounded Model Checker will
%% create a SAT problem of size $O(lk)$.  Conversely, a SAT problem
%% of size $s$ can be converted trivially into a loop-free \newC program
%% of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
%% which means it can be decided fairly efficiently for many practical
%% instances.

\subsection{Encoding into \newC}
%% To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
%% \sigma$ and $P$ in \newC, then ensure that we can express the validity of the
%% synthesis formula as a safety property of the resulting \newC program.

@@

@@

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.65, transform shape, node distance=2cm, auto]
\tikzstyle{decision} = [diamond, draw, fill=blue!20, 
    text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
\tikzstyle{block} = [rectangle, draw, fill=blue!20, 
    text width=5em, text centered, rounded corners, minimum height=4em]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
    minimum height=2em]

 \node [decision] (synthsucc) {{\sc Synth} succeeds?};

 \node [decision, below of=synthsucc, node distance=2.5cm] (verif) {{\sc Verif} succeeds?};
 \node [decision, right of=verif] (ck) {$c < l$?};

 \node [block, below of=verif, node distance=3cm]   (done) {Done!};
 \node [decision, left of=verif] (verifw) {{\sc Verif} succeeds for small words?};

 \node [block, below of=ck, node distance=2.5cm] (incc) {$c := c+1$};
 \node [block, right of=incc, node distance=2cm] (incl) {$c := 0$\\ $l := l+1$};

 \node [decision, below of=verifw, node distance=3cm] (gen) {Extend succeeds?};
 \node [block, left of=verifw, node distance=3cm] (iterate) {Parameters unchanged};

 \node [block, left of=gen, node distance=3cm] (incw) {$w := w+1$};

 \path [line] (synthsucc) -- node [left] {Yes} (verif);
 \path [line] (synthsucc) -| node [above, near start] {No} (ck);

 \path [line] (verif) -- node [left] {Yes} (done);
 \path [line] (verif) -- node [above, near start] {No} (verifw);

 \path [line] (ck) -- node [left] {Yes} (incc);
 \path [line] (ck) -| node  [above, near start]  {No} (incl);

 \path [line] (verifw) -- node [right] {Yes} (gen);
 \path [line] (verifw) -- node [above] {No} (iterate);

 \path [line] (gen) -- node [below] {Yes} (done);
 \path [line] (gen) -- node [below] {No} (incw);

 %\path [dotted, line] (iterate.west) |- (synthsucc);
\end{tikzpicture}

 \caption{Decision tree for increasing parameters of $\mathcal{L}$.}
 \label{fig:paramsflow}

>>>>

In this section we present a solver for the synthesis fragment. An
important characteristic of our solver is that is models bit accurate
semantics. This enables the precise treatment of bit-wise operations and
arithmetic modulo fixed widths.
% 
%We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
%have a program synthesiser kicking around, we present the design of one synthesiser
%that can be used to solve the problem.
%We start by recalling the synthesis formula:
% $$\exists P, x . \forall y . \sigma(P, x, y)$$
We start by recalling the synthesis fragment:
 \[
  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
 \]

@@

\subsection{\newC}
\label{sec:logic}
In order to obtain bit-level accuracy, the logic we use to express the synthesis formula is a subset of C that we call \newC.  
The characteristic property of a
\newC  program is that safety can be decided for
it using a single query to a Bounded Model Checker.  A \newC program is
just a C program with the following syntactic restrictions:
 all loops in the program must have a constant bound;
 all recursion in the program must be limited to a constant depth;
 all arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
Additionally, \newC programs may use nondeterministic values, assumptions
and arbitrary-width types.

The proofs found by our solver are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
%written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
\todo{update the figure and make it look more like a grammar}

As the quantifier free first-order fragment of the logic is efficiently decidable and
$P$ is expressible as a ground term of the logic, we can use
Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
check the validity of the synthesis formula.

The core of the CEGIS algorithm is
the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
detailed in Algorithm~\ref{fig:abstract-refinement-code}.  
The algorithm is divided into two
components: {\sc synth} and {\sc verif}. For a finite set of test
inputs that we track, {\sc synth} synthesises a candidate program that is correct for those tests.
Subsequently, {\sc Verif} verifies the candidate program by searching for an input on
which it does not satisfy the specification. If no such input
can be found, the candidate program is correct.  Otherwise, the new input is added
to the set of tracked inputs.  
%If the input set is finite, this procedure is guaranteed to terminate.

@@

%% To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
%% \sigma$ and $P$ in \newC, then ensure that we can express the validity of the
%% synthesis formula as a safety property of the resulting \newC program.

@@

---
We require the user to supply a program specification written in C, but then
ask nothing more of her.  We parametrise the space of programs in such a way
that it can be explored automatically, rather than asking a human for hints.
Our program encoding generates an $\exists
\forall$ formula that is \emph{linear} in the length of the shortest program
satisfying the user's specification.  This formula is then checked for satisfiability
using the CEGIS algorithm.
---

Our tool is the first we are aware of that is able to effectively
synthesise floating-point programs.  We demonstrate this by
synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
a specification.
------

@@

%% \begin{figure}[t]
%% \centering
%% \begin{tikzpicture}[scale=0.65, transform shape, node distance=2cm, auto]
%% \tikzstyle{decision} = [diamond, draw, fill=blue!20, 
%%     text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
%% \tikzstyle{block} = [rectangle, draw, fill=blue!20, 
%%     text width=5em, text centered, rounded corners, minimum height=4em]
%% \tikzstyle{line} = [draw, -latex']
%% \tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
%%     minimum height=2em]

%%  \node [decision] (synthsucc) {{\sc Synth} succeeds?};

%%  \node [decision, below of=synthsucc, node distance=2.5cm] (verif) {{\sc Verif} succeeds?};
%%  \node [decision, right of=verif] (ck) {$c < l$?};

%%  \node [block, below of=verif, node distance=3cm]   (done) {Done!};
%%  \node [decision, left of=verif] (verifw) {{\sc Verif} succeeds for small words?};

%%  \node [block, below of=ck, node distance=2.5cm] (incc) {$c := c+1$};
%%  \node [block, right of=incc, node distance=2cm] (incl) {$c := 0$\\ $l := l+1$};

%%  \node [decision, below of=verifw, node distance=3cm] (gen) {Extend succeeds?};
%%  \node [block, left of=verifw, node distance=3cm] (iterate) {Parameters unchanged};

%%  \node [block, left of=gen, node distance=3cm] (incw) {$w := w+1$};

%%  \path [line] (synthsucc) -- node [left] {Yes} (verif);
%%  \path [line] (synthsucc) -| node [above, near start] {No} (ck);

%%  \path [line] (verif) -- node [left] {Yes} (done);
%%  \path [line] (verif) -- node [above, near start] {No} (verifw);

%%  \path [line] (ck) -- node [left] {Yes} (incc);
%%  \path [line] (ck) -| node  [above, near start]  {No} (incl);

%%  \path [line] (verifw) -- node [right] {Yes} (gen);
%%  \path [line] (verifw) -- node [above] {No} (iterate);

%%  \path [line] (gen) -- node [below] {Yes} (done);
%%  \path [line] (gen) -- node [below] {No} (incw);

%%  %\path [dotted, line] (iterate.west) |- (synthsucc);
%% \end{tikzpicture}

%%  \caption{Decision tree for increasing parameters of $\mathcal{L}$.}
%%  \label{fig:paramsflow}

%% \end{figure}

### 2014-07-02 papers/termination/paper.tex
working on the kalashnikov section

>>>>

\todo{do something with this?}
Since each loop is bounded by a constant, and each recursive function call is
limited to a constant depth, a \newC program necessarily terminates and in
fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
a Bounded Model Checker with an unrolling bound of $k$ will be a complete
decision procedure for the safety of the program.  For a \newC program of
size $l$ and with largest loop bound~$k$, a Bounded Model Checker will
create a SAT problem of size $O(lk)$.  Conversely, a SAT problem
of size $s$ can be converted trivially into a loop-free \newC program
of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
which means it can be decided fairly efficiently for many practical
instances.
